sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-12-09 doc/paper.Rmd
Edits

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017] (`?pc_matern`) can constrain the spatial range and variance parameters.

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary difference is the user interface.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA may find inlabru approachable.
Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets (e.g., REFs).
INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.

>>>>

sdmTMB models can include penalized likelihoods by assigning priors to model parameters (`?sdmTMBpriors`).
These priors may be useful in cases where estimation is difficult because of identifiability issues or relatively flat likelihood surfaces, or to impart prior information or achieve regularization.
Following other recent SPDE implementations in TMB [@breivik2021; @osgood-zimmerman2021], penalized complexity ('PC') priors [@simpson2017; @fuglstad2019] (`?pc_matern`) can constrain the spatial range and variance parameters.

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table 1), we suggest the primary difference is the user interface.
We think users familiar with `stats::glm()`, lme4, or glmmTMB will find sdmTMB most approachable.
Users familiar with INLA may find inlabru approachable.
Users already familiar with mgcv can fit adapt mgcv to fit similar models with custom code [@miller2019] and INLA/inlabru and mgcv are also general purpose modelling packages.
VAST is the sole option for fitting some multivariate models;
alternatively, because VAST focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemoral GLMMs may find sdmTMB more straightforward to use.
Users looking for calculation with uncertainty of derived variables such as area-weighted population indexes may favour sdmTMB or VAST (although such quantities can be post hoc derived with other packages).
Speed wise, sdmTMB (and VAST) may be fastest in some cases (e.g., random effect normalization) or in situations up to mid-resolution meshes (~500 knots), which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @english2021].
INLA/inlabru, or spaMM for spatial models, scale the best with very high resolution meshes.
